Repeatability of Pre-Flashover Fire Patterns on Gypsum Wall Board

### 2015-03-23 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Madrzy_Thesis: latex cleanup

\usepackage{graphicx}
\frenchspacing

@@

\abstract{
Uncontrolled fires result in loss of life, property and can create an adverse economic impact on a community.  The investigation of fires provides a means to identify the cause of the fire in order to develop a knowledge base that could enable the elimination of that cause and thus reduce the losses from fires.  However, as the result of the review of several arson cases and a review by the U.S. National Academy of Sciences, questions about the lack of the use of science in the practice of fire investigation were raised.  Specifically the National Academy of Sciences indicated that ``... research is needed on the natural variability of burn patterns...''
This study addresses that need in two ways:
\begin{enumerate}
\item Examining the repeatability of several small fire sources and the fire patterns that were generated by those fires
\item Examining the capability of using numerical models to simulate the fires and the resulting fire patterns based on input data collected from engineering reference sources, and bench scale and full scale fire experiments conducted in this study.
\end{enumerate}
This paper highlights the many uncertainties involved in what appear to be simple fire experiments.  A variety of uncertainties related to the composition of the fuels, measurement methods, and analysis techniques just to list a few of the areas considered.  The objective of the research was to examine the repeatability of pre-flashover fire patterns generated from exposure to short duration (300 s maximum) well characterized fires.  Three different fuels were used: natural gas, gasoline, and polyurethane foam.  Each fuel had a similar top surface area.  The heat release rate data showed that the variability was greater for more complex fuels.  While the variation in peak heat release rate with the natural gas was similar to the expanded uncertainty of the measurement system, 11~\%, the variation in the peak heat release rates of the gasoline and the polyurethane foam increased due to increased uncertainties in the burning behavior of the fuels.  The patterns generated by the polyurethane foam fire, had more variability than the natural gas and the gasoline fire patterns.  This is likely due to the greater variability of the burning of the solid fuel and the lower heat release rate.  The maximum fire pattern heights generated from the natural gas and gasoline fires were shown to have uncertainties of 18~\% or less based on a Type A statistical analysis with 95~\% confidence limits.  The comparison of the fire pattern heights and the mean flame height demonstrated that the ``'steady state'' natural gas fires exhibited the highest level of agreement and polyurethane foam fueled fire exhibited the least agreement, with the gasoline fueled fires in between.

@@

The source fire characterization experiments were conducted in the NIST Large Fire Research Laboratory’s 3~m by 3~m hood oxygen consumption calorimeter. This apparatus follows the methodology of ASTM~E 2067-08, {\em Standard Practice for Full-Scale Oxygen Consumption Calorimetry Fire Tests}~\cite{ASTM_E2067}. The fires were centered under the hood of the calorimeter and positioned in an open steel frame which provided support for the thermocouples and heat flux sensors. A load cell was positioned as a support for the fuels.  All of the data was recorded at intervals of 1~s on a computer based data acquisition system.

@@

On each fuel’s averaged heat release rate curve, a bar representing the 95~\% confidence level ($2\sigma$), based on Type A evaluation of the uncertainty is provided~\cite{Taylor:1994}.  A Type A evaluation of the uncertainty is any statistical analysis of the data from a series of observations.  In this case the evaluation assumed a normal distribution of the data such that doubling the calculated standard deviation (2$\sigma$) would yield a confidence level of approximately 95~\%.

@@

\begin{figure}
  \includegraphics[width=2.7in]{../Figures/Fig15a}
  \includegraphics[width=2.7in]{../Figures/Fig15b} \\
  \caption[Photographs of the gasoline fire]{Photograph showing a gasoline fueled experiment with a flame height of 0.65 m (left) and a photograph taken approximately 0.2 s later with a flame height of 1.0 m (right).}
  \label{Gasoline_Photos}
\end{figure}

@@

Examples of a set of TIFF images for each fuel are shown in Figures~\ref{flame_height_comp_FHNG80_1} through \ref{flame_height_comp_FHPUF_1}.   Each figure has 56 sequential images from the video segment.  The images total almost 2~s of actual burn time.  Notice how the flame height cycles up and down in each of the sequences.  These images provide additional information and continuity of the motion of the flame so that the analysis is less likely to miss and extreme (minimum or maximum) position when compared to the still photographs.
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/flame_height_comp_FHNG80}\\
	\caption[Example of sequential video frames sampled from a natural gas fueled experiment]{Example of sequential video frames sampled from a natural gas fueled experiment.  In this set of images, which spans almost two seconds, the visual continuous flame height fluctuates from approximately 0.6 m to 1.1 m.}
	\label{flame_height_comp_FHNG80_1}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/flame_height_comp_FHGAS}\\
	\caption[Example of sequential video frames sampled from a gasoline fueled experiment]{Example of sequential video frames sampled from a gasoline fueled experiment.  In this set of images which spans almost two seconds the visual continuous flame height fluctuates from approximately 0.6 m to 1.1 m.}
	\label{flame_height_comp_FHGAS_1}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/flame_height_comp_FHPUF}\\
	\caption[Example of sequential video frames sampled from a polyurethane foam fueled experiment]{Example of sequential video frames sampled from a polyurethane foam fueled experiment.  In this set of images which spans almost two seconds the continuous visual flame height fluctuates from 0.4 m to 0.9 m.}
	\label{flame_height_comp_FHPUF_1}
\end{figure}

@@

Now that the characteristics of the fires have been measured, the next step is to measure the thermal impact that the fires have on a target.  In these experiments the target is a non-combustible wall. The non-combustible material was chosen to eliminate any contribution from the wall to the fire.  The fires were initially placed at predetermined, fixed positions away from the wall.  In the remainder of the report the term burner will be considered synonymous with the pan for the liquid fuel or the block of solid fuel. Experiments were conducted with the closest edge of the burner located in the following positions from the wall: two burner widths (0.610 m), one burner width (0.305 m), one half burner width (0.150 m), and the touching the wall.

@@

Figure~\ref{IW_SC_TWGAS} shows the temperatures of the array of thermocouples located along the vertical centerline of the gasoline fueled burner.  The beads of the thermocouples were installed in the wall and were flush with the wall surface. In each graph, the average temperature of the replicate experiments, for each thermocouple position is shown.  Moving from 0.2 m above the top surface of the burner up to 2.2 m above the top of the burner surface in increments of 0.2 m.

@@

Figure~\ref{IW_HFC_TWGAS} shows the temperatures of the array of total heat flux gauges located along the vertical centerline of the gasoline fueled burner.  The sensing surface of the gauges were installed flush with the wall surface and exposed to the fire. In each graph, the average temperature of the replicate experiments, for each total heat flux position is shown.  Moving from 0.2 m above the top surface of the burner up to 1.2 m above the top of the burner surface in increments of 0.2 m.

@@

The vertical radiant panel is used as documented in ASTM~E162 and normally provides a measure of downward, opposed-flow flame spread~\cite{ASTM_E162}.  The apparatus measures the surface flammability of materials using a pre-mixed air-gas-fired radiant panel that is 305~mm by 457~mm.

@@

The rear wall of the compartment was constructed in the same manner for a given set of experiments. The burner was centered on the middle panel of the rear wall.  As in the fire source characterization experiments the top of the fuel in each case was positioned 7.5~cm above the floor of the compartment.  Each of the painted gypsum board panels that were used for a fire pattern had a 5 cm grid drawn on it to assist with the determination of the geometry of the fire pattern.

@@

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|} {\textwidth}
		\hline
	Fuel (number of tests) (\%)   &       Wall Type	            &   Max. Height (m)	            & Max. Width (m)    &  Height at Max. Width (m)    &  Area (m $^2$)  \\ \hline
		Natural Gas (10)    &   Open 		& 0.74 	$\pm$ 0.12 (16~\%)   &	0.24 $\pm$ 0.06 (25~\%) 	& 0.41 $\pm$ 0.07 (17~\%)	& 0.15 $\pm$ 0.05 (33~\%) 	\\
		Natural Gas (10)    &   Closed 		& 0.74	$\pm$ 0.06 (8~\%)    &	0.24 $\pm$ 0.02 (7~\%) 		& 0.43 $\pm$ 0.03 (7~\%)  	& 0.14 $\pm$ 0.01 (6~\%) 	\\
		Natural Gas (10)    &   Insulated 	& 0.71	$\pm$ 0.10 (14~\%)   &	0.23 $\pm$ 0.04 (17~\%) 	& 0.42 $\pm$ 0.07 (17~\%) 	& 0.15 $\pm$ 0.02 (14~\%)  	\\
		Gasoline (12)       &   Open    	& 0.83	$\pm$ 0.15 (18~\%)   &	0.28 $\pm$ 0.09 (32~\%) 	& 0.44 $\pm$ 0.14 (32~\%) 	& 0.17 $\pm$ 0.04 (24~\%)  	\\
		Gasoline (10)       &   Closed  	& 0.80	$\pm$ 0.10 (13~\%)   &	0.29 $\pm$ 0.04 (14~\%) 	& 0.44 $\pm$ 0.06 (14~\%) 	& 0.17 $\pm$ 0.04 (24~\%)  	\\
		Gasoline (10)       &   Insulated	& 0.82 	$\pm$ 0.09 (11~\%)   &	0.25 $\pm$ 0.08 (32~\%) 	& 0.40 $\pm$ 0.13 (32~\%) 	& 0.13 $\pm$ 0.04 (30~\%)  	\\

		\hline
	\end{tabular}
	\caption[Comparison of fire pattern dimensions as a function of wall construction type]{Comparison of fire pattern dimensions as a function of wall construction type. Values presented with 95\% confidence limits.}
	\label{tab:Fire_Pattern_Dimensions_Wall_Construction}
\end{table}

>>>>

\usepackage{graphicx}
\usepackage{rotating}
\frenchspacing

@@

\abstract{
Uncontrolled fires result in loss of life, property and can create an adverse economic impact on a community.  The investigation of fires provides a means to identify the cause of the fire in order to develop a knowledge base that could enable the elimination of that cause and thus reduce the losses from fires.  However, as the result of the review of several arson cases and a review by the U.S. National Academy of Sciences, questions about the lack of the use of science in the practice of fire investigation were raised.  Specifically the National Academy of Sciences indicated that ``... research is needed on the natural variability of burn patterns...''
This study addresses that need in two ways:
\begin{enumerate}
\item Examining the repeatability of several small fire sources and the fire patterns that were generated by those fires
\item Examining the capability of using numerical models to simulate the fires and the resulting fire patterns based on input data collected from engineering reference sources, and bench scale and full scale fire experiments conducted in this study.
\end{enumerate}
This paper highlights the many uncertainties involved in what appear to be simple fire experiments.  A variety of uncertainties related to the composition of the fuels, measurement methods, and analysis techniques just to list a few of the areas considered.  The objective of the research was to examine the repeatability of pre-flashover fire patterns generated from exposure to short duration (300 s maximum) well characterized fires.  Three different fuels were used: natural gas, gasoline, and polyurethane foam.  Each fuel had a similar top surface area.  The heat release rate data showed that the variability was greater for more complex fuels.  While the variation in peak heat release rate with the natural gas was similar to the expanded uncertainty of the measurement system, 11~\%, the variation in the peak heat release rates of the gasoline and the polyurethane foam increased due to increased uncertainties in the burning behavior of the fuels.  The patterns generated by the polyurethane foam fire, had more variability than the natural gas and the gasoline fire patterns.  This is likely due to the greater variability of the burning of the solid fuel and the lower heat release rate.  The maximum fire pattern heights generated from the natural gas and gasoline fires were shown to have uncertainties of 18~\% or less based on a Type A statistical analysis with 95~\% confidence limits.  The comparison of the fire pattern heights and the mean flame height demonstrated that the ``steady state'' natural gas fires exhibited the highest level of agreement and polyurethane foam fueled fire exhibited the least agreement, with the gasoline fueled fires in between.

@@

The source fire characterization experiments were conducted in the NIST Large Fire Research Laboratory’s 3~m by 3~m hood oxygen consumption calorimeter. This apparatus follows the methodology of ASTM~E 2067-08, {\em Standard Practice for Full-Scale Oxygen Consumption Calorimetry Fire Tests} ~\cite{ASTM_E2067}. The fires were centered under the hood of the calorimeter and positioned in an open steel frame which provided support for the thermocouples and heat flux sensors. A load cell was positioned as a support for the fuels.  All of the data was recorded at intervals of 1~s on a computer based data acquisition system.

@@

On each fuel’s averaged heat release rate curve, a bar representing the 95~\% confidence level (2$\sigma$), based on Type A evaluation of the uncertainty is provided~\cite{Taylor:1994}.  A Type A evaluation of the uncertainty is any statistical analysis of the data from a series of observations.  In this case the evaluation assumed a normal distribution of the data such that doubling the calculated standard deviation (2$\sigma$) would yield a confidence level of approximately 95~\%.

@@

\begin{figure}
  \includegraphics[width=2.7in]{../Figures/Fig15a}
  \includegraphics[width=2.7in]{../Figures/Fig15b} \\
  \caption[Photographs of the gasoline fire]{Photograph showing a gasoline fueled experiment with a flame height of 0.65~m (left) and a photograph taken approximately 0.2~s later with a flame height of 1.0~m (right).}
  \label{Gasoline_Photos}
\end{figure}

@@

Examples of a set of TIFF images for each fuel are shown in Figures~\ref{flame_height_comp_FHNG80_1} through \ref{flame_height_comp_FHPUF_1}.   Each figure has 56 sequential images from the video segment.  The images total almost 2~s of actual burn time.  Notice how the flame height cycles up and down in each of the sequences.  These images provide additional information and continuity of the motion of the flame so that the analysis is less likely to miss and extreme (minimum or maximum) position when compared to the still photographs.
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/flame_height_comp_FHNG80}\\
	\caption[Example of sequential video frames sampled from a natural gas fueled experiment]{Example of sequential video frames sampled from a natural gas fueled experiment.  In this set of images, which spans almost two seconds, the visual continuous flame height fluctuates from approximately 0.6~m to 1.1~m.}
	\label{flame_height_comp_FHNG80_1}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/flame_height_comp_FHGAS}\\
	\caption[Example of sequential video frames sampled from a gasoline fueled experiment]{Example of sequential video frames sampled from a gasoline fueled experiment.  In this set of images which spans almost two seconds the visual continuous flame height fluctuates from approximately 0.6~m to 1.1~m.}
	\label{flame_height_comp_FHGAS_1}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/flame_height_comp_FHPUF}\\
	\caption[Example of sequential video frames sampled from a polyurethane foam fueled experiment]{Example of sequential video frames sampled from a polyurethane foam fueled experiment.  In this set of images which spans almost two seconds the continuous visual flame height fluctuates from 0.4~m to 0.9~m.}
	\label{flame_height_comp_FHPUF_1}
\end{figure}

@@

Now that the characteristics of the fires have been measured, the next step is to measure the thermal impact that the fires have on a target.  In these experiments the target is a non-combustible wall. The non-combustible material was chosen to eliminate any contribution from the wall to the fire.  The fires were initially placed at predetermined, fixed positions away from the wall.  In the remainder of the report the term burner will be considered synonymous with the pan for the liquid fuel or the block of solid fuel. Experiments were conducted with the closest edge of the burner located in the following positions from the wall: two burner widths (0.610~m), one burner width (0.305~m), one half burner width (0.150~m), and the touching the wall.

@@

Figure~\ref{IW_SC_TWGAS} shows the temperatures of the array of thermocouples located along the vertical centerline of the gasoline fueled burner.  The beads of the thermocouples were installed in the wall and were flush with the wall surface. In each graph, the average temperature of the replicate experiments, for each thermocouple position is shown.  Moving from 0.2~m above the top surface of the burner up to 2.2~m above the top of the burner surface in increments of 0.2~m.

@@

Figure~\ref{IW_HFC_TWGAS} shows the temperatures of the array of total heat flux gauges located along the vertical centerline of the gasoline fueled burner.  The sensing surface of the gauges were installed flush with the wall surface and exposed to the fire. In each graph, the average temperature of the replicate experiments, for each total heat flux position is shown.  Moving from 0.2~m above the top surface of the burner up to 1.2~m above the top of the burner surface in increments of 0.2~m.

@@

The vertical radiant panel is used as documented in ASTM~E162 and normally provides a measure of downward, opposed-flow flame spread~\cite{ASTM_E162}.  The apparatus measures the surface flammability of materials using a pre-mixed air-gas-fired radiant panel that is 305~mm × 457~mm.

@@

The rear wall of the compartment was constructed in the same manner for a given set of experiments. The burner was centered on the middle panel of the rear wall.  As in the fire source characterization experiments the top of the fuel in each case was positioned 7.5~cm above the floor of the compartment.  Each of the painted gypsum board panels that were used for a fire pattern had a 5~cm grid drawn on it to assist with the determination of the geometry of the fire pattern.

@@

\begin{sidewaystable}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
	Fuel (\# of tests)      &   Wall Type   &   Max. Height (m)	         & Max. Width (m)               &  Height at Max. Width (m) &  Area (m $^2$)  \\ \hline
		Natural Gas (10)    &   Open 		& 0.74 	$\pm$ 0.12 (16~\%)   &	0.24 $\pm$ 0.06 (25~\%) 	& 0.41 $\pm$ 0.07 (17~\%)	& 0.15 $\pm$ 0.05 (33~\%) 	\\
		Natural Gas (10)    &   Closed 		& 0.74	$\pm$ 0.06 (8~\%)    &	0.24 $\pm$ 0.02 (7~\%) 		& 0.43 $\pm$ 0.03 (7~\%)  	& 0.14 $\pm$ 0.01 (6~\%) 	\\
		Natural Gas (10)    &   Insulated 	& 0.71	$\pm$ 0.10 (14~\%)   &	0.23 $\pm$ 0.04 (17~\%) 	& 0.42 $\pm$ 0.07 (17~\%) 	& 0.15 $\pm$ 0.02 (14~\%)  	\\
		Gasoline (12)       &   Open    	& 0.83	$\pm$ 0.15 (18~\%)   &	0.28 $\pm$ 0.09 (32~\%) 	& 0.44 $\pm$ 0.14 (32~\%) 	& 0.17 $\pm$ 0.04 (24~\%)  	\\
		Gasoline (10)       &   Closed  	& 0.80	$\pm$ 0.10 (13~\%)   &	0.29 $\pm$ 0.04 (14~\%) 	& 0.44 $\pm$ 0.06 (14~\%) 	& 0.17 $\pm$ 0.04 (24~\%)  	\\
		Gasoline (10)       &   Insulated	& 0.82 	$\pm$ 0.09 (11~\%)   &	0.25 $\pm$ 0.08 (32~\%) 	& 0.40 $\pm$ 0.13 (32~\%) 	& 0.13 $\pm$ 0.04 (30~\%)  	\\
		\hline
	\end{tabular}
	\caption[Comparison of fire pattern dimensions as a function of wall construction type]{Comparison of fire pattern dimensions as a function of wall construction type. Values presented with 95~\% confidence limits.}
	\label{tab:Fire_Pattern_Dimensions_Wall_Construction}
\end{sidewaystable}
